\vspace{-\baselineskip} Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-2\baselineskip}

### 2017-05-08 papers/paper1/ms.tex
Corrected indexing in equations (21)-(24)

The relationship between the \celeriteterm\ model and semi-separable linear
algebra was first recognized by Ambikasaran \citep{Ambikasaran:2015}.
A rank-$R$ semi-separable matrix is a matrix $K$ where the elements can be
written as
\begin{eqnarray}
K_{n,m} = \left\{\begin{array}{ll}
    \sum_{r=1}^R U_{n,r}\,V_{m,r} & \mathrm{if}\,m<n \\
    A_{n,n} & \mathrm{if}\,m=n \\
    \sum_{r=1}^R U_{m,r}\,V_{n,r} & \mathrm{otherwise}
\end{array}\right.
\end{eqnarray}
for some $N \times R$-dimensional matrices $U$ and $V$, and an $N \times
N$-dimensional diagonal matrix $A$.
This definition can be equivalently written as
\begin{eqnarray}\eqlabel{semi-sep}
K = A + \mathrm{tril}(U\,V^\T) + \mathrm{triu}(V\,U^\T)
\end{eqnarray}
where the $\mathrm{tril}$ function extracts the strict lower triangular
component of its argument and $\mathrm{triu}$ is the equivalent strict upper
triangular operation.
For the \celeriteterm\ model described in the previous section, the rank of
the semi-separable matrix is $R = 2\,J$ and the components of the relevant
matrices are
\begin{eqnarray}
U_{n,2\,j} &=& a_j\,e^{-c_j\,t_n}\,\cos(d_j\,t_n) +
    b_j\,e^{-c_j\,t_n}\,\sin(d_j\,t_n) \\
U_{n,2\,j+1} &=& a_j\,e^{-c_j\,t_n}\,\sin(d_j\,t_n) -
    b_j\,e^{-c_j\,t_n}\,\cos(d_j\,t_n) \\
V_{m,2\,j} &=& e^{c_j\,t_m}\,\cos(d_j\,t_m) \\
V_{m,2\,j+1} &=& e^{c_j\,t_m}\,\sin(d_j\,t_m) \\
A_{n,n} &=& \sigma_n^2 + \sum_{j=1}^J a_j \quad.
\end{eqnarray}

@@

These special cases demonstrate that the stochastically-driven simple harmonic
oscillator provides a physically motivated model that is flexible enough to
describe a wide range of stellar variations and we return to give some
specific examples in \sect{examples-real}.
Low $Q \approx 1$ can capture granulation noise and high $Q \gg 1$ is a good
model for asteroseismic oscillations.
In practice, we take a sum over oscillators with different values of $Q$,
$S_0$, and $\omega_0$ to give a sufficient accounting of the power spectrum
stellar time series.
Since this kernel is exactly described by the exponential kernel, the
likelihood (\eqalt{gp-likelihood}) can be evaluated for a time series with $N$
measurements in $\mathcal{O}(N)$ operations using the \celeriteterm\ method
described in the previous section.

>>>>

The relationship between the \celeriteterm\ model and semi-separable linear
algebra was first recognized by Ambikasaran \citep{Ambikasaran:2015}.
A rank-$R$ semi-separable matrix is a matrix $K$ where the elements can be
written as
\begin{eqnarray}
K_{n,m} = \left\{\begin{array}{ll}
    \sum_{r=1}^R U_{n,r}\,V_{m,r} & \mathrm{if}\,m<n \\
    A_{n,n} & \mathrm{if}\,m=n \\
    \sum_{r=1}^R U_{m,r}\,V_{n,r} & \mathrm{otherwise}
\end{array}\right.
\end{eqnarray}
for some $N \times R$-dimensional matrices $U$ and $V$, and an $N \times
N$-dimensional diagonal matrix $A$.
This definition can be equivalently written as
\begin{eqnarray}\eqlabel{semi-sep}
K = A + \mathrm{tril}(U\,V^\T) + \mathrm{triu}(V\,U^\T)
\end{eqnarray}
where the $\mathrm{tril}$ function extracts the strict lower triangular
component of its argument and $\mathrm{triu}$ is the equivalent strict upper
triangular operation.
For the \celeriteterm\ model described in the previous section, the rank of
the semi-separable matrix is $R = 2\,J$ and the components of the relevant
matrices are
\begin{eqnarray}
U_{n,2\,j-1} &=& a_j\,e^{-c_j\,t_n}\,\cos(d_j\,t_n) +
    b_j\,e^{-c_j\,t_n}\,\sin(d_j\,t_n) \\
U_{n,2\,j} &=& a_j\,e^{-c_j\,t_n}\,\sin(d_j\,t_n) -
    b_j\,e^{-c_j\,t_n}\,\cos(d_j\,t_n) \\
V_{m,2\,j-1} &=& e^{c_j\,t_m}\,\cos(d_j\,t_m) \\
V_{m,2\,j} &=& e^{c_j\,t_m}\,\sin(d_j\,t_m) \\
A_{n,n} &=& \sigma_n^2 + \sum_{j=1}^J a_j \quad.
\end{eqnarray}

@@

These special cases demonstrate that the stochastically-driven simple harmonic
oscillator provides a physically motivated model that is flexible enough to
describe a wide range of stellar variations and we return to give some
specific examples in \sect{examples-real}.
Low $Q \approx 1$ can capture granulation noise and high $Q \gg 1$ is a good
model for asteroseismic oscillations.
In practice, we take a sum over oscillators with different values of $Q$,
$S_0$, and $\omega_0$ to give a sufficient accounting of the power spectrum
of stellar time series.
Since this kernel is exactly described by the exponential kernel, the
likelihood (\eqalt{gp-likelihood}) can be evaluated for a time series with $N$
measurements in $\mathcal{O}(N)$ operations using the \celeriteterm\ method
described in the previous section.
